Thought of as the quantum counterpart to classical random walks, quantum walks~\cite{QW-overview} provide an interesting technique in algorithmic design, 
with  applications in unstructured search, graph algorithmics and communication protocols.

Differently from the classical case, where the  walker's next move follows the result of some sort of random choice, in a quantum setting evolution typically 
proceeds in an equally weighed superposition of possible moves through the iteration of a unitary operator, without resorting to intermediate measurements. 
This results in a very rich dynamics, in which the design of the evolution operator, and even seemingly
innocent differences  in its phase  and in the initial state, determine complex `walking patterns'  
which differ greatly both from each other and from the classical setting.

The relevance of quantum walks as a tool for algorithmic design justifies both a better understanding of their behaviour and the optimisation of their implementation, namely to increase resilience to decoherence phenomena. This  paper resorts to the ZX-calculus~\cite{CoeckeD08,ZX-overview,CoeckeHKW22} for such a purpose. The ZX-calculus is a diagrammatic language for reasoning about linear maps between qubits. It is based on generators --- the \emph{spiders} --- that generalise rotations over the $Z$ and $X$ basis, and sets of rewrite rules whose completeness means that  equality between linear maps can be proven diagrammatically. This allows for program transformation in quantum software engineering, guided by powerful  circuit optimisation strategies for e.g. T-count reduction and gate compilation.

Optimisation of quantum circuits can be seen as a \emph{reconfiguration} process. Indeed, as shown in the following sections, the interpretation of such circuits as ZX-diagrams provides a flexible description of quantum computations graphically. Then, the rules of the ZX-calculus guide through a simplification strategy which corresponds to sequences of graph transformations. Finally, the reconfigured  circuit is extracted from the transformed graph. The process is illustrated here in a closed setting. However, it extends smoothly to the dynamic case as required by quantum algorithms which reconfigure themselves as a result of the (classical) evaluation of measurement results.  This is particularly relevant in the context of variational algorithms~\cite{Cer21} which are core to current techniques in quantum machine learning~\cite{Dun16}.



The exercise reported here focuses on a recent, very general variant of discrete-time quantum walks --- the \emph{staggered}
 model~\cite{Portugal_2017,PortugalSFG16} --- which, avoiding the use of a coin to direct the walker evolution, explores the underlying graph structure to build an evolution operator based on local unitaries induced by adjacent vertices.  The model is reviewed in Section 2. Then, in Section 3, its standard circuit implementation is translated and rewritten in ZX, supported by  the \textit{PyZX} tool~\cite{pyzx}. This process leads in Section 4 to the identification of a diagrammatic pattern providing an interesting approximation to, and in some cases,  more efficient version, of the underlying evolution operator.  



